---
title: "Natural Organic Matter Data Notebook"
author: "Bea Meluch"
date: "2024-07-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(jsonlite)
library(heatmaply)
```

# Define functions to interact with API

## Define a general API call funtion to nmdc-runtime

This function provides a general-purpose way to make an API request to NMDC's runtime API. Note that this function will only return the first page of results. The function's input includes the name of the collection to access (e.g. biosample_set), the filter to be performed, the maximum page size, and a list of the fields to be retrieved. It returns the metadata as a dataframe.

```{r def-results-first}
get_first_page_results <- function(collection, filter, max_page_size, fields) {
  og_url <- paste0(
      'https://api.microbiomedata.org/nmdcschema/', 
      collection, '?&filter=', filter, '&max_page_size=', max_page_size, '&projection=', fields
      )
  
  response <- jsonlite::fromJSON(URLencode(og_url, repeated = TRUE))
  
  return(response)
}
```

## Define an nmdc-runtime API call function to include pagination

The get_next_results function uses the get_first_page_results function, defined above, to retrieve the rest of the results from a call with multiple pages. It takes the same inputs as the get_first_page_results function above: the name of the collection to be retrieved, the filter string, the maximum page size, and a list of the fields to be returned. This function returns the results as a single dataframe (can be nested). It uses the next_page_token key in each page of results to retrieve the following page.

```{r def-results-next}
get_next_results <- function(collection, filter_text, max_page_size, fields) {
  initial_data <- get_first_page_results(collection, filter_text, max_page_size, fields)
  results_df <- initial_data$resources
  
  if (!is.null(initial_data$next_page_token)) {
    next_page_token <- initial_data$next_page_token
    
    while (TRUE) {
      url <- paste0('https://api.microbiomedata.org/nmdcschema/', collection, '?&filter=', filter_text, '&max_page_size=', max_page_size, '&page_token=', next_page_token, '&projection=', fields)
      response <- jsonlite::fromJSON(URLencode(url, repeated = TRUE))

      results_df <- results_df %>% bind_rows(response$resources)
      next_page_token <- response$next_page_token
      
      if (is.null(next_page_token)) {
        break
      }
    }
  }
  
  return(results_df)
}
```



## Define an API request function that uses a list of ids to filter on

This function constructs a different type of API request that takes a list of ids or similar (e.g. `biosample` ids as retrieved above). The `id_field` input is a string of the name of the id field name (e.g. `id` or `has_output`), the name of the new collection to be queried, the name of the field to match the previous ids on in the new collection, and a list of the fields to be returned.

```{r def-results-id}
get_results_by_id <- function(collection, match_id_field, id_list, fields, max_id = 50) {
    # collection: the name of the collection to query
    # match_id_field: the field in the new collection to match to the id_list
    # id_list: a list of ids to filter on
    # fields: a list of fields to return
    # max_id: the maximum number of ids to include in a single query
    
    # If id_list is longer than max_id, split it into chunks of max_id
    if (length(id_list) > max_id) {
        id_list <- split(id_list, ceiling(seq_along(id_list)/max_id))
    } else {
        id_list <- list(id_list)
    }
    
    output <- list()
    for (i in 1:length(id_list)) {
        # Cast as a character vector and add double quotes around each ID
        mongo_id_string <- as.character(id_list[[i]]) %>%
            paste0('"', ., '"') %>%
            paste(collapse = ', ')
        
        # Create the filter string
        filter = paste0('{"', match_id_field, '": {"$in": [', mongo_id_string, ']}}')
        
        # Get the data
        output[[i]] = get_next_results(
            collection = collection,
            filter = filter,
            max_page_size = max_id*3, #assumes that there are no more than 3 records per query
            fields = fields
        )
    }
    output_df <- bind_rows(output)
}
```

# Query 

## get all NOM data

```{r}
# Get data objects using get_next_results function
nom_dobj_df <- get_next_results(
    collection = "data_object_set", 
    filter_text = '{"data_object_type": {"$regex": "FT ICR-MS Analysis Results"}}', 
    max_page_size = 100, 
    fields = "id,name,url,was_generated_by,md5_checksum"
    )
```


## get all nom analysis results by id 

```{r}
nom_analysis_df <- get_results_by_id(
    collection = "nom_analysis_activity_set",
    match_id_field = "has_output",
    id_list = nom_dobj_df$id,
    fields = "id,has_input,has_output",
    max_id = 20
) %>%
  mutate(has_input = unlist(has_input),
         has_output = unlist(has_output))

```

## get all dataobjects of raw NOM data by id
```{r}

nom_raw_df <- get_results_by_id(
    collection = "data_object_set",
    match_id_field = "id",
    id_list = nom_analysis_df$has_input,
    fields = "id,was_generated_by,name,md5_checksum",
    max_id = 20
)

# uh oh, there are dobj with no was_generated_by, summarize
# nom_raw_df %>%
#   filter(is.na(was_generated_by)) %>%
#   mutate(name_prefix = str_match(name, "^(output: [A-Za-z]+)_")[,2]) %>%
#   count(name_prefix)

```

## get all omicsprocessing records by id
```{r}

nom_omprc_df <- get_results_by_id(
    collection = "omics_processing_set",
    match_id_field = "has_output",
    id_list = nom_raw_df$id,
    fields = "id,has_input,has_output",
    max_id = 20
) %>%
  mutate(has_input = unlist(has_input),
         has_output = unlist(has_output))

```

## get all biosample records by id
```{r}

nom_biosample_df <- get_results_by_id(
    collection = "biosample_set",
    match_id_field = "id",
    id_list = nom_omprc_df$has_input,
    fields = "id,name,part_of,analysis_type,geo_loc_name,env_medium",
    #fields = "",
    max_id = 20
)

```

## combine records into one data frame

```{r}

nom_objects_joined <- nom_biosample_df %>%
  rename_with(~ paste0("bsm.", .x)) %>%
  full_join(rename_with(nom_omprc_df, ~ paste0("omprc.", .x)), 
            by = join_by(bsm.id == omprc.has_input),
            keep = TRUE) %>%
  full_join(rename_with(nom_raw_df, ~ paste0("raw.", .x)), 
            by = join_by(omprc.has_output == raw.id),
            keep = TRUE) %>%
  full_join(rename_with(nom_analysis_df, ~ paste0("analysis.", .x)), 
            by = join_by(raw.id == analysis.has_input),
            keep = TRUE) %>%
  full_join(rename_with(nom_dobj_df, ~ paste0("dobj.", .x)), 
            by = join_by(analysis.has_output== dobj.id),
            keep = TRUE) %>%
  
  # clean up: remove duplicate rows produced by joins
  distinct() %>%
  
  # flatten list-columns
  unnest(where(is.data.frame), names_sep = ".", keep_empty = TRUE) %>%
  unnest(where(is.data.frame), names_sep = ".", keep_empty = TRUE) %>%
  mutate(across(where(is.list), as.character)) %>%
  
  # remove duplicate processed data objects
  filter(!duplicated(dobj.md5_checksum))


```



# Label sample types
```{r sample-types}

nom_objects_joined_labeled <- nom_objects_joined %>%


  mutate(bsm.sample_type_manual = case_when(
    bsm.env_medium.term.id %in% c("UBERON:0035118", "ENVO:00002003") ~ "host",
    str_detect(bsm.name, "hizosphere") ~ "plant",
    str_detect(bsm.env_medium.term.id, "PO:") ~ "plant",
    bsm.env_medium.term.id %in% c("ENVO:00001998", "ENVO:00005774", "ENVO:00002258", "ENVO:00005778", "ENVO:00005802", 
                                  "ENVO:02000059", "ENVO:00005741", "ENVO:01001841", "ENVO:00003082", "ENVO:00005800", "ENVO:01000018") ~ "soil",
    bsm.env_medium.term.id %in% c("ENVO:00002007", "ENVO:01000017", "ENVO:03605004", "ENVO:03605005") ~ "sediment",
    bsm.env_medium.term.id %in% c("ENVO:00002042", "ENVO:00002011", "ENVO:00002194", "ENVO:01001869", "ENVO:04000007") ~ "water",
    bsm.env_medium.term.id %in% c("ENVO:01000349", "ENVO:01001001", "ENVO:03605005") ~ "plant",
    str_detect(bsm.env_medium.term.name, "soil") ~ "soil",
    str_detect(bsm.env_medium.has_raw_value, "soil") ~ "soil",
    str_detect(bsm.env_medium.term.name, "water") ~ "water",
    bsm.part_of == "nmdc:sty-11-pzmd0x14" ~ "sediment"
  ))

```

# Pull processed data

```{r}
# Pull processed data files using URLs from processed data dobj
# processed_data_csvs <- data.frame(data_object_url = nom_objects_joined_labeled$dobj.url) %>%
#   filter(!is.na(data_object_url)) %>%
#   mutate(csv_nested = lapply(data_object_url, read.csv))
# 
# saveRDS(processed_data_csvs, "processed_data_csvs.RDS")

processed_data_csvs <- readRDS("processed_data_csvs.RDS")

```

## Quality metrics for processed data
```{r qc}
# Calculate some statistics for the processed data, plot to assess quality
processed_data_qc <- processed_data_csvs %>%
  mutate(median_confidence  = vapply(csv_nested, function(x) median(x$Confidence.Score, na.rm = TRUE), 
                                     FUN.VALUE = 1.0, USE.NAMES = FALSE),
         
         unassigned_count   = vapply(csv_nested, function(x) sum((x$Molecular.Formula == "")), 
                                     FUN.VALUE = 1, USE.NAMES = FALSE),
         
         assigned_count     = vapply(csv_nested, function(x) sum((x$Molecular.Formula != "")), 
                                     FUN.VALUE = 1, USE.NAMES = FALSE),
         
         assigned_perc      = assigned_count / (unassigned_count + assigned_count),
         
         molecular_formulae = lapply(csv_nested, function(x) unique(x$Molecular.Formula))) %>%
  
  select(-csv_nested)



ggplot(processed_data_qc) +
  geom_histogram(aes(x = median_confidence), bins = 20) +
  ggtitle("Median Confidence Scores for Processed Data Files")

ggplot(processed_data_qc) +
  geom_histogram(aes(x = assigned_perc)) +
  ggtitle("Percent Assigned Peaks for Processed Data Files")

```

```{r pairwise-shared-formulae}







```


```{r formula-heatmap}

# filter out low quality samples to reduce computational time

samples_for_heatmap <- processed_data_qc %>%
  filter(assigned_perc > 0.3) %>%
  left_join(select(nom_objects_joined_labeled, dobj.url, bsm.id, bsm.sample_type_manual, bsm.part_of), 
            by = join_by(data_object_url == dobj.url)) %>%
  group_by(bsm.sample_type_manual) %>%
  slice_sample(n = 33)






all_formulae <- samples_for_heatmap$molecular_formulae %>%
  unlist() %>%
  unique()

all_formulae_filtered <- samples_for_heatmap$molecular_formulae %>%
  unlist() %>%
  table() %>%
  as.data.frame() %>%
  setNames(c("Formula", "Freq")) %>%
  filter(Formula != "") %>%
  #filter(Freq >= nrow(samples_for_heatmap)/2) %>%
  filter(Freq > 5) %>%
  mutate(Formula = as.character(Formula)) %>%
  pull(Formula)
  

# initialize presence absence matrix
# rows are samples
# columns are molecular formulae
holder <- matrix(, nrow = nrow(samples_for_heatmap), ncol = length(all_formulae_filtered))

for (i in 1:nrow(samples_for_heatmap)) {
  holder[i,] <- ifelse(all_formulae_filtered %in% samples_for_heatmap$molecular_formulae[[i]], 
                       1, 0)
}

rownames(holder) <- samples_for_heatmap$bsm.id
colnames(holder) <- all_formulae_filtered


sample_type_color_vec <- case_when(
  samples_for_heatmap$bsm.sample_type_manual == "soil" ~ "saddlebrown",
  samples_for_heatmap$bsm.sample_type_manual == "sediment" ~ "slategray",
  samples_for_heatmap$bsm.sample_type_manual == "water" ~ "blue")

study_color_vec <- case_when(
  samples_for_heatmap$bsm.part_of == "nmdc:sty-11-aygzgv51" ~ "red",
  samples_for_heatmap$bsm.part_of == "nmdc:sty-11-5tgfr349" ~ "orange",
  samples_for_heatmap$bsm.part_of == "nmdc:sty-11-28tm5d36" ~ "yellow",
  samples_for_heatmap$bsm.part_of == "nmdc:sty-11-db67n062" ~ "green",
  samples_for_heatmap$bsm.part_of == "nmdc:sty-11-dcqce727" ~ "blue",
  samples_for_heatmap$bsm.part_of == "nmdc:sty-11-8xdqsn54" ~ "purple",
  samples_for_heatmap$bsm.part_of == "nmdc:sty-11-33fbta56" ~ "pink"
  )


heatmap(holder, Colv = NA, RowSideColors = sample_type_color_vec, labRow = FALSE, labCol = FALSE, scale = "none")

# heatmap(holder) there is way too much data for this to work, try something else


heatmaply(holder, Colv = NA, RowSideColors = cbind(study_color_vec, sample_type_color_vec), showticklabels = FALSE, hide_colorbar = TRUE)




```


```{r van-krevelen}

# extract oc and hc ratios for the same samples as used in the heatmap

samples_for_heatmap %>%
  select(data_object_url, bsm.id, bsm.sample_type_manual, bsm.part_of) %>%
  left_join(processed_data_csvs) %>%
  mutate(HC = lapply(csv_nested, function(x) x$H.C),
         OC = lapply(csv_nested, function(x) x$O.C)) %>%
  select(-csv_nested) %>%
  unnest() %>%

  ggplot(aes(x = OC, y = HC, color = bsm.sample_type_manual)) +
    geom_point(alpha = 0.1, size = 0.5) +
    facet_wrap(~bsm.sample_type_manual) +
    theme(legend.position = "none") +
    coord_fixed()


```